### 2022-09-03 main/chapter_searches.tex
some frequentist text

Coin tosses, for example, can be imagined as random.
Earth's human population could toss a few billion coins to precisely
establish a probability of landing heads from the ratio of
$n$ (number of heads) to $N$ (number of tosses).
Now suppose you are given a red envelope containing a coin;
which way up is the coin in the envelope?
You could guess or bet on the answer, and that decision might involve a tacitly
assumed probability.
A frequency, however, would only be permitted if the envelope had been packaged
randomly.
Was the coin first tossed to randomize its face, or placed deliberately?
It is unclear.

Tom Stoppard's play `Rosencrantz and Guildenstern Are Dead'~\cite{
stoppard1967rosencrantz
}
opens with a clearer example ---
the titular characters toss a coin ninety-two times, all landing heads
(by ``un-, sub- or supernatural forces'').
They then play a game with a third player, landing eight more heads.
Will the next toss be heads or tails?
The answer is not random --- it is burned in print onto paper, so no frequency
can be assigned.
Were subjective probabilities permitted, you could predict the result, and that
prediction would be subjectively informed by it being the one-hundred-and-first
toss, our cultural entanglement with round numbers in base-10, and your reading
of these words.
Frequentist theory, however, forbids such subjective probabilities of
non-random unknowns, so is left seeking alternatives.

@@

Modern frequentist methods are largely attributable to the authors
R. A. Fisher (for likelihood, $p$-values, and estimation criteria), and
J. Neyman and E. S. Pearson
(for statistical hypothesis testing and confidence intervals).
Together these do not form a harmonious framework;
the two groups have emphatic disagreements in print, but together they
do cover many practices that survive in modern science and explain important
features of the methods we apply to LHC data, which will be introduced in
Section~\ref{sec:searches_practice}.

@@

\TODO{
Nayman-Pearson:
Seek good test statistics --
likelihood ratio intuitive for two-hypothesis comparison
prove it optimal in a reasonable sense
idea of long-run mostly not wrong
recede to maximum likelihood ratios
~\cite{neymanpearson1928max}
``for which hypothesis A is more probable and those for which it is less
probable (sic!)''
unbiased (unbiassed) test differs from unbiasedness of estimator
``Then the two corresponding densities at the sample point
are (Al) 6-42 x 10-20, (A2) 5 18*x 1O-4 and the ratio of the likelihood of Hypothesis
Al to- that of A2 is 1P24 x 10-16. This ratio confirms the common-sense judgment
that A2 is far more plausible than A1''
\^ Another posterior argument --- this likelihood ratio in in fact a change
in plausibilities. For example, if A2 is an inconsistent, self-contradictory
theory, it is totally implausible, and not more plausible despite predicting
these data better.
``The statistician can
balance the numerical verdict of likelihood against the vaguer expectations de-
rived from `a priori' considerations, and it must be left to his judgment to
at what point the evidence in favour of alternative hypotheses becomes so con-
vincing that Hypothesis A must be rejected.''
Introduce maximum likelihood ratio as interesting without claiming it to be
necessarily best.
Neyman-Pearson lemma --- reproduce Fig 5?
~\cite{neymanpearson19233lemma}
}

\begin{figure}[tp]
\centering
\includegraphics[width=0.95\textwidth]{figures/searches_np_curves_crop.png}
\caption[
Tail areas of prior data distributions from two alternative hypotheses,
reproduced from Neyman and Pearson
]{%
Tail areas of prior data distributions from two alternative hypotheses,
reproduced from Neyman and Pearson~\cite{neymanpearson19233lemma}.
}
\label{fig:searches_np_tails}
\end{figure}

>>>>

Coin tosses, for example, can be imagined as random.
Earth's human population could toss a few billion coins to precisely
establish a probability of landing heads from the ratio of
$n$ (number of heads) to $N$ (number of tosses).
Now suppose you are given a red envelope containing a coin;
which way up is the coin in the envelope?
You could guess or bet on the answer, and that decision might involve a tacitly
assumed probability.
A frequency, however, would only be permitted if the envelope had been packaged
randomly.
Was the coin first tossed to randomize its face, or placed deliberately?
You do not know.

Tom Stoppard's play `Rosencrantz and Guildenstern Are Dead'~\cite{
stoppard1967rosencrantz
}
opens with a clearer example ---
the titular characters toss a coin ninety-two times, all landing heads.
They then play a game with a third player, landing eight more heads.
Will the next toss be heads or tails?
The answer is not random --- it is burned onto paper in print, so no frequency
can be assigned.
Were subjective probabilities permitted, you could predict the result, and that
prediction would be subjectively informed by it being the one-hundred-and-first
toss, our cultural entanglement with round numbers in base-10, and your reading
of these words.
Frequentist theory, however, forbids such subjective probabilities of
non-random unknowns, so is left to search for alternative modes of inference.

@@

Modern frequentist methods are largely attributable to work by
Ronald A. Fisher (for likelihood, $p$-values, and estimation criteria~\cite{
fisher1921probable,
fisher1922estimators,
fisher1925smrw,
fisher1956statistical
}),
and Jerzy Neyman and Egon S. Pearson
(for statistical hypothesis testing and confidence intervals~\cite{
neymanpearson1933lemma,
neymanpearson1928max
})
in the 1920s and 30s.
Together, the Fisher and Neyman-Pearson methods do not form a harmonious
framework, but comprise a loose collection of ideas and practices that survive
in modern science, and explain important features of the methods we apply to
LHC data, which Section~\ref{sec:searches_practice} develops further.

Some literature names this Fisher/Neyman-Pearson assemblage
``classical'' statistics~\cite{lehmann2011fisher, Feldman:1997qc}, but that is
simply inaccurate for a 20th century divergence from 18th century
ideas~\cite{bayes1763lii, laplace1774stigler}.
Jaynes instead names them ``orthodox'~\cite{jaynes2003probability},
which may be accurate in some contexts, but alternative
(heretical?) ideas have taken hold under modern LHC authorities ---
the orthodoxy has moved on.
We therefore stick to the name `frequentist', as it is a firm common factor.

% likelihood
Fisher coined the term `likelihood'~\cite{fisher1921probable}, in the usage
we have employed, for the probability of observed data given varied
contexts\footnote{%
Fisher used the term `likelihood' for the function exploring
contexts and `probability' for the distribution over
outcomes~\cite{fisher1925smrw}.
We choose the alternative word `prior' for his `probability' since the
likelihood is also a probability, just not a probability \emph{distribution}.
},
and deserves credit for popularising its use in interpreting and communicating
data.
% figure here TODO
% likelihood quotes
Fisher forcefully argued against the use of Bayes' rule and prior
probabilities, stating for example that
\begin{singlespacing}
\begin{displayquote}
We can know nothing of the probability of hypotheses or hypothetical
quantities.
On the other hand we may ascertain the likelihood of hypotheses
and hypothetical quantities by calculation from observations~\cite{
fisher1921probable
}
\end{displayquote}
\end{singlespacing}

% p-values

% estimators --- not unbiased!

% savage likelihood principle

@@

\TODO{
Nayman-Pearson:
Seek good test statistics --
likelihood ratio intuitive for two-hypothesis comparison
prove it optimal in a reasonable sense
idea of long-run mostly not wrong
recede to maximum likelihood ratios
~\cite{neymanpearson1928max}
``for which hypothesis A is more probable and those for which it is less
probable (sic!)''
unbiased (unbiassed) test differs from unbiasedness of estimator
``Then the two corresponding densities at the sample point
are (Al) 6-42 x 10-20, (A2) 5 18*x 1O-4 and the ratio of the likelihood of Hypothesis
Al to- that of A2 is 1P24 x 10-16. This ratio confirms the common-sense judgment
that A2 is far more plausible than A1''
\^ Another posterior argument --- this likelihood ratio in in fact a change
in plausibilities. For example, if A2 is an inconsistent, self-contradictory
theory, it is totally implausible, and not more plausible despite predicting
these data better.
``The statistician can
balance the numerical verdict of likelihood against the vaguer expectations de-
rived from `a priori' considerations, and it must be left to his judgment to
at what point the evidence in favour of alternative hypotheses becomes so con-
vincing that Hypothesis A must be rejected.''
Introduce maximum likelihood ratio as interesting without claiming it to be
necessarily best.
Neyman-Pearson lemma --- reproduce Fig 5?
~\cite{neymanpearson1933lemma}
}

\begin{figure}[tp]
\centering
\includegraphics[width=0.95\textwidth]{figures/searches_np_curves_crop.png}
\caption[
Tail areas of prior data distributions from two alternative hypotheses,
reproduced from Neyman and Pearson
]{%
Tail areas of prior data distributions from two alternative hypotheses,
reproduced from Neyman and Pearson~\cite{neymanpearson1933lemma}.
}
\label{fig:searches_np_tails}
\end{figure}
